---
title: "Learning R"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Objects in **R**

Almost anything can be stored as an object in **R**.  Here, we save the operation of `1 + 1` to an object called `x` and then print that object.

```{r r_object, echo=TRUE}

x <- 1 + 1
x

```

The `<-` symbol tells **R** you want to assign whatever is on the right of the sybol to the object on the left of the symbol.  In this example, we are assigning the value of `1 + 1` to the object `x`.  An important point is that **R** is evaluating the expression `1 + 1` and storing the value of the expression.  It doesn't actually store `1 + 1`.  Also, by default, **R** doesn't display objects unless explicitly requested.   


### Exercise 

Write the R code required to add two plus two and store it as an object called `y` and print the object:

```{r two_plus_two, exercise=TRUE}

```

```{r two_plus_two-solution}
y <- 2 + 2
y
```


### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Reading in data

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

