---
title: "RMarkdown"
author: "Matthew Cox-Martin"
date: "August 10, 2017"
output: html_document
---

# This is the exercise from the "Advanced R" tutorial.  

## Setup

The code below sets up the R environment by loading the `tidyverse` package.

```{r setup, echo=TRUE}
options(scipen=999, digits = 5, width = 88)
options(qwraps2_markup = "markdown")

options(stringAsFactors = F)
options(dplyr.print_max = 10)
options(dplyr.pring_min = 5)

knitr::opts_chunk$set(cache   = FALSE,
               echo    = TRUE,
               results = "hide",
               message = FALSE,
               warning = FALSE) 

options(repos = c(CRAN = "https://cran.rstudio.com"))

library(tidyverse)

```

## Import

The code below imports the data using `readr::read_csv`.  The `csv` stands for "Comma Seperate Values".

```{r import, echo=TRUE}

data <- read_csv("https://raw.githubusercontent.com/vtmacox/teachr/master/data.csv", col_names = TRUE)

```

## Data Cleaning

The code below assigns the `med` variable as a factor and provides labels.  It also recodes values of `9` as missing for the `anxiety` variable.

```{r data_setup, echo=TRUE}
data$med <- factor(data$med,
                   levels = c(0,1),
                   labels = c("Not on medication", "On Medication"))

data$chol <- as.numeric(data$chol)

data$anxiety <- if_else(data$anxiety == 9, NA_integer_, data$anxiety)

# Or you could do it this way
# data$anxiety <- ifelse(data$anxiety == 9, NA, data$anxiety)

```

## Descriptive Stats

The code below shows two methods of creating descriptive stats.  The first uses `psych::describe`.  This prints out useful infromation such as mean, median, range, min, max, skew, and kurtosis.  However, if you have factor variables, it does not output frequences.  The second method uses functions from the `qwraps2` package which creates tables that can be displayed in markdown documents.  

```{r desc_stats, echo=TRUE}

psych::describe(data)

tab1 <- 
  with(data,
       list( "Hours of Exercise" = qwraps2::tab_summary(exer),
             "Servings of Fruits and Vegetables" = qwraps2::tab_summary(fv),
             "Age" = qwraps2::tab_summary(age),
             "Anxiety" = qwraps2::tab_summary(anxiety),
             "Cholesterol Level" = qwraps2::tab_summary(chol),
             "On Cholesterol Medication" = qwraps2::tab_summary(med))
)

```

## Display descriptive stats

```{r disp_stat, results="asis", echo=TRUE}

outtab1 <- qwraps2::summary_table(data, tab1)
outtab1

```


## Plots

Here we are using ggplot2.

```{r plots, echo=TRUE}

ggplot(data=data, aes(x = med, y= as.numeric(chol), fill = med)) +
  stat_summary(fun.y = "mean", geom = "bar") + 
  theme(axis.title.x = element_blank()) + 
  ylab("Cholesterol (mg/dL)") +  guides(fill=FALSE) + 
  ggtitle("Cholesterol Levels by Medication Status")

ggplot(data=data, aes(x=age, y=chol)) +
  geom_point(shape = 1) + 
  ylab("Cholesterol (mg/dL)") + 
  geom_smooth(method = lm, se = FALSE) +
  ggtitle("Relationship between Cholesterol and Age") +
  xlab("Age (years)")

```

## Regression Model

```{r reg_mod, results="markdown"}
# Linear Regression
mod1 <- lm(chol ~ exer + fv + age + anxiety + med, data = data)

summary(mod1)
```

## Other display options

```{r reg_mod, results="asis"}
# Method 1 for output
knitr::kable(summary(mod1)$coef, digits = 5)

# Method 2 for output
pander::pander(summary(mod1))

# Method 3 for output (may have to install xtable package)
xtable::print.xtable(xtable::xtable(summary(mod1)), type = "html")
```